---
title: "| RESEARCH PROTOCOL\n| \n| Estimating the comparative risk of kidney failure associated with intravitreal anti-vascular endothelial growth factor exposure in patients with blinding diseases \n"
fontsize: 12pt
geometry: margin=1in
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../docs") })
output:
  bookdown::html_document2:
    df_print: paged
    toc: yes
    toc_depth: 2
    toc_float: yes
    number_sections: yes
    number_tables: yes
    css: style.css
  word_document:
    reference_docx: ohdsi-protocol-style.docx
  bookdown::pdf_document2:
    keep_tex: yes
    latex_engine: xelatex
    md_extensions: +raw_attribute
    number_sections: yes
    includes:
      before_body: title.tex
header-includes:
- \usepackage[numbers,sort&compress]{natbib}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{caption}
- \usepackage{rotating}
- \usepackage{multirow}
- \usepackage{mwe,tikz}
- \usepackage[percent]{overpic}
- \usepackage{enumitem}
- \usepackage{hyperref}
- \newcolumntype{P}[1]{>{\raggedright\arraybackslash}p{#1}}
- \newcommand{\footerDate}{`r params$date`}
- \input{header.tex}
longtable: yes
mainfont: Arial
bibliography: Protocol.bib
params:
  date: '28-Mar-2023'
  version: 0.0.1
subtitle: 'Version: `r params$version`'
link-citations: yes
csl: bmj.csl
---

```{r setup, include=FALSE}
# library(LegendT2dm)
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
options(kableExtra.latex.load_packages = FALSE)
library(kableExtra)
library(dplyr)
options(readr.show_col_types = FALSE)
options(knitr.kable.NA = "")
options(knitr.table.format = function() {
  if (knitr::is_latex_output()) {
    "latex"
  } else if (knitr::is_html_output()) {
    "html"
  } else {
    "pipe"
  }
})
latex_table_font_size <- 8
```

# List of Abbreviations

```{r abbreviations, echo=FALSE}
abbreviations <- readr::read_delim(col_names = FALSE, delim = ",", trim_ws = TRUE, file = "
Anti-VEGF, anti-vascular endothelial growth factor 
DME, diabetic macular edema 
DR, diabetic retinopathy 
AMD, age-related macular degeneration 
VO, vein occlusion 
CKD, chronic kidney disease 
ICD, International Classification of Diseases 
CPT, Current Procedural Terminology 
OHDSI, Observational Health Data Sciences and Informatics 
OMOP, Observational Medical Outcomes Partnership 
CDM, common data model 
PS, propensity score 
EHR, electronic health record 
IRB, Institutional Review Board 
JHM, Johns Hopkins Medicine 
ARVO, Association for Research in Vision and Ophthalmology 
")

tab <- kable(abbreviations, col.names = NULL, linesep = "", booktabs = TRUE)

if (knitr::is_latex_output()) {
  tab %>% kable_styling(latex_options = c("striped", "hold_position"), 
                        font_size = latex_table_font_size)
} else {
  tab %>% kable_styling(bootstrap_options = "striped")
}
```

\clearpage

# Responsible Parties

## Investigators

```{r parties, echo=FALSE}
parties <- readr::read_delim(col_names = TRUE, delim = ";", trim_ws = TRUE, file = "
  Investigator; Institution/Affiliation
  Cindy Cai, MD *; Wilmer Eye Institute, Johns Hopkins Hospital, Baltimore, MD, USA
")

tab <- kable(parties, booktabs = TRUE, linesep = "") %>%
  column_spec(1, width = "10em") %>%
  column_spec(2, width = "35em") %>%
  footnote(general = "* Principal Investigator", general_title = "")

if (knitr::is_latex_output()) {
  tab %>% kable_styling(latex_options = c("striped", "hold_position"),
                        font_size = latex_table_font_size)
} else {
  tab %>% kable_styling(bootstrap_options = "striped")
}
```

## Disclosures

This study is undertaken within Observational Health Data Sciences and Informatics (OHDSI), an open collaboration.

\clearpage

# Abstract

Background: The potential adverse kidney effects of the commonly used intravitreal anti-vascular endothelial growth factor (anti-VEGF) medications are poorly understood.
Purpose: To estimate the risk of kidney failure after intravitreal anti-VEGF medications across the Observational Health Data Sciences and Informatics (OHDSI) network, comparing ranibizumab, aflibercept, and bevacizumab.
Design: Retrospective analysis of administrative claims and electronic health record (EHR) data in the OHDSI network.
Subjects: Patients with blinding diseases including diabetic macular edema (DME), diabetic retinopathy (DR), age-related macular degeneration (AMD), and vein occlusion (VO) who receive ≥3 monthly anti-VEGF will be included. Patients with pre-existing kidney failure will be excluded.
Methods: Patients receiving ranibizumab will be compared to propensity score matched patients receiving aflibercept, and patients receiving bevacizumab. Propensity score models will be constructed on baseline characteristics such as demographics (age, sex, race/ethnicity), comorbidities (Charlson Comorbidity Index), and receipt of any nephrotoxic medications during the observation window. Cox models will be used to compare the time to kidney failure between patients who receive ranibizumab with patients who receive aflibercept, ranibizumab with bevacizumab, and bevacizumab with aflibercept.
Conclusions: Knowing the relative risk for kidney failure between the three commonly used anti-VEGF medications will allow ophthalmologists to select safer treatments.


\clearpage

# Amendments and Updates

```{r amendments, echo=FALSE}
amendments <- readr::read_delim(col_names = TRUE, delim = ";", trim_ws = TRUE, file = "
  Number; Date; Section of study protocol; Amendment or update; Reason
")

tab <- kable(amendments, booktabs = TRUE, linesep = "")

if (knitr::is_latex_output()) {
  tab %>% kable_styling(latex_options = c("striped", "hold_position"), font_size = latex_table_font_size)
} else {
  tab %>% kable_styling(bootstrap_options = "striped")
}
```

# Milestones

```{r dates, echo=FALSE}
dates <- readr::read_delim(col_names = TRUE, delim = ";", trim_ws = TRUE, file = "
•	Identify data partners
•	Data partners to share data diagnostics results
•	Complete candidate phenotypes
•	Data partners to share phenotype diagnostics results
•	Finalize analytic package
•	Data partners to run analytic package
•	Data partners to share study diagnostic results
•	Finalize results from data partners
•	Manuscript drafting and writing
•	Submit abstract: OHDSI Global Symposium (May 2023), ARVO (Jan 2024)
")

tab <- kable(dates, booktabs = TRUE, linesep = "")

if (knitr::is_latex_output()) {
  tab %>% kable_styling(latex_options = c("striped", "hold_position"), font_size = latex_table_font_size)
} else {
  tab %>% kable_styling(bootstrap_options = "striped")
}
```

# Rationale and Background
Intravitreal anti-vascular endothelial growth factor (anti-VEGF) medications have revolutionized the treatment of blinding diseases.1,2 Nearly 20 million intravitreal injections were given globally in 2016 to treat conditions such as diabetic macular edema (DME), diabetic retinopathy (DR), age-related macular degeneration (AMD), and vein occlusion (VO).1 When given systemically, anti-VEGF medications have adverse kidney effects (e.g., proteinuria, acute kidney injury).3–5 Patients at high risk for chronic kidney disease (CKD), for example, those ≥60 years and those with diabetes, may be preferentially harmed by the adverse kidney effects of intravitreal anti-VEGF.6 CKD can progress to kidney failure, which is a major source of morbidity and mortality for the patient and is a financial burden to society.6,7

When anti-VEGF medications are given intravitreally, there is systemic absorption of the drug that is greatest with bevacizumb and aflibercept and lowest with ranibizumab.8,9 However, evidence about the subsequent kidney effect is conflicting. Several case reports document kidney damage after intravitreal anti-VEGF.10–12 On the other hand, retrospective cohort studies have not identified treatment-associated declines in kidney function, or estimated glomerular filtration rate (eGFR).13,14 A secondary analysis from a randomized trial also did not show differences in rates of kidney damage.15 However, these studies were either limited to a single institution, had small sample sizes (<600 patients), only included patients without CKD, and did not evaluate the severe phenotype of kidney failure. To resolve this controversy, a comprehensive study is needed to evaluate the kidney effects of intravitreal anti-VEGF. In populations at high risk for CKD, the treating ophthalmologist can choose alternate forms of therapy (e.g., focal laser for DME), or the intravitreal anti-VEGF with the least systemic absorption (e.g., ranibizumab).16–18


# Study Objectives
The overall objective of the proposal is to estimate the comparative risk of kidney failure associated with intravitreal anti-vascular endothelial growth factor exposure in patients with blinding diseases. The rationale is that knowing the comparative risk for kidney failure among the three commonly used anti-VEGF medications will allow the ophthalmologist to select the safest treatments, particularly for populations at high risk for adverse kidney effects. To address our hypothesis, we propose the following specific aims.
1.	Amongst people with blinding diseases (DR/DME, AMD, and VO), does exposure to ranibizumab increase the risk of kidney failure, relative to aflibercept?
2.	Amongst people with blinding diseases (DR/DME, AMD, and VO), does exposure to ranibizumab increase the risk of kidney failure, relative to bevacizumab?
3.	Amongst people with blinding diseases (DR/DME, AMD, and VO), does exposure to bevacizumab increase the risk of kidney failure, relative to aflibercept?

# Research Methods

## Study Design/Data Sources
This will be a retrospective study of patients with blinding diseases (DR/DME, AMD, and VO) comparing the risk of kidney failure after exposure to ranibizumab relative to aflibercept, ranibizumab to bevacizumab, and bevacizumab to aflibercept. Data sources will be administrative claims or electronic health record (EHR) data across the OHDSI network.

## Study Population
Patients ≥18 years who have received ≥3 monthly intravitreal anti-VEGF injections with ranibizumab, aflibercept, or bevacizumab for a diagnosis of DME/DR, AMD, and VO will be included in the study.16,17,19–22 Cohort entry will begin with the initiation of the anti-VEGF medications. Patients with pre-existing kidney failure will be excluded. (Figure 1, Figure 2) For the primary analysis, only patients who are on monotherapy (e.g., only ranibizumab, only aflibercept, only bevacizumab) will be included.

## Exposure Comparators
The exposure is receipt of ≥3 monthly intravitreal anti-VEGF medications (whether ranibizumab, aflibercept, or bevacizumab).

## Outcomes {#outcomes}
The outcome is time from anti-VEGF initiation to kidney failure, as defined by previously validated OMOP phenotypes using ICD and CPT codes, during the observation window.23

## Covariates
For databases that provide sufficient sample sizes (>= 1,000) in each target and comparator cohorts, we will deploy the large-scale propensity score method that adjust for a comprehensive range of baseline characteristics.24 For smaller databases, we will select a smaller number chosen based on expert opinion: demographics (age, sex, race/ethnicity), ophthalmic indication for anti-VEGF (DME/DR, AMD, or VO), ophthalmic comorbidities (e.g., glaucoma, cataracts, other retinal diseases), medical comorbidities (Charlson Comorbidity Index), and receipt of any nephrotoxic medications during the observation window (e.g., aminoglycosides).25

# Analysis Plan
## Calculation of time-at risk
The time-at-risk for patients who undergo anti-VEGF will be calculated as the periods of time when patients undergo monthly or bimonthly injections. In other words, a patient is considered no longer at risk once there is a >90 days-gap in anti-VEGF administration. Anti-VEGF need typically decreases over time; however, some patients require lifelong VEGF suppression and the proportion of patients who require lifelong VEGF is likely different depending on the disease indication (e.g., DME/DR, versus AMD, versus VO).26–30

## Model Specification/Analysis
Propensity score methods will be used to match/stratify patients in each target and comparator cohorts. Propensity score matched/stratified Cox proportional hazards models will be used to estimate the risk of kidney failure.31

## Sensitivity Analyses
Various sensitivity analyses will be performed. 1) Multiple definitions of the cohort entry criteria will be used, specifically how 3 “monthly” anti-VEGF medications will be defined. For example, one definition could include a first anti-VEGF, followed by another anti-VEGF 4-8 weeks after the first one, and another anti-VEGF 8-16 weeks after the first one. Alternatively, the intervals of 0, 4-8, and 8-20 will be used. The least strict criteria will be the designation of ≥3 injections in the first 24 weeks. 2) The anti-VEGF injection treatment schedules are expected to differ among patients with the three ophthalmic diagnoses (DME/DR, versus AMD, versus VO). Therefore, we will estimate the risk differences within each diagnosis in addition to the joint analysis.

# Sample Size and Study Power {#sample-size}

# Study Diagnostics
1.	Preference score distributions to evaluate empirical equipoise and population generalizability
2.	Extensive patient characteristics to evaluate cohort balance before and after PS-adjustment
3.	Negative control calibration plots to assess residual bias
4.	Kaplan-Meier plots to examine hazard ratio proportionality assumptions


# Strengths and Limitations {#strengths-limitations}
When executed, this will be the largest study examining the association between intravitreal anti-VEGF medications and kidney failure. Other strengths of this study include the diversity of the sample (from multiple institutions, multiple countries, and a range of underlying kidney impairment at baseline).

Limitations are mainly due to the retrospective nature of the study which might have confounding despite propensity score matching and stratified analyses.

# Protection of Human Subjects
The JHM IRB has already approved this project (IRB00356149). Each participating institution will seek IRB approval for this study.

# Management and Reporting of Adverse Events and Adverse Reactions
JHM will serve as the coordinating center for this OHDSI network study. All findings and adverse events/reactions will be reported to them.

# Plans for Disseminating and Communicating Study Results
The results will be shared and discussed among the study participants and broader OHDSI community during the weekly Tuesday community calls throughout the SOS Challenge. This work will be presented at conferences (OHDSI Global Symposium and/or ARVO) and published as a manuscript.

\clearpage

# References {-}

1. Martin DF. Evolution of Intravitreal Therapy for Retinal Diseases—From CMV to CNV: The LXXIV Edward Jackson Memorial Lecture. Am J Ophthalmol. 2018;191:xli-lviii. doi:10.1016/j.ajo.2017.12.019
2. Collaborators G 2019 B and VI, Adelson JD, Bourne RRA, et al. Causes of blindness and vision impairment in 2020 and trends over 30 years, and prevalence of avoidable blindness in relation to VISION 2020: the Right to Sight: an analysis for the Global Burden of Disease Study. Lancet Global Heal. 2021;9(2):e144-e160. doi:10.1016/s2214-109x(20)30489-7
3. Hanna RM, Barsoum M, Arman F, Selamet U, Hasnain H, Kurtz I. Nephrotoxicity Induced by Intravitreal Vascular Endothelial Growth Factor (VEGF) inhibitors: Emerging Evidence. Kidney Int. 2019;96(3):572-580. doi:10.1016/j.kint.2019.02.042
4. Gurevich F, Perazella MA. Renal Effects of Anti-angiogenesis Therapy: Update for the Internist. Am J Medicine. 2009;122(4):322-328. doi:10.1016/j.amjmed.2008.11.025
5. Izzedine H, Escudier B, Lhomme C, et al. Kidney Diseases Associated With Anti-Vascular Endothelial Growth Factor (VEGF). Medicine. 2014;93(24):333-339. doi:10.1097/md.0000000000000207
6. Levey AS, Atkins R, Coresh J, et al. Chronic kidney disease as a global public health problem: Approaches and initiatives – a position statement from Kidney Disease Improving Global Outcomes. Kidney Int. 2007;72(3):247-259. doi:10.1038/sj.ki.5002343
7. Collaboration GCKD, Bikbov B, Purcell CA, et al. Global, regional, and national burden of chronic kidney disease, 1990–2017: a systematic analysis for the Global Burden of Disease Study 2017. Lancet. 2020;395(10225):709-733. doi:10.1016/s0140-6736(20)30045-3
8. Avery RL, Castellarin AA, Steinle NC, et al. Systemic pharmacokinetics following intravitreal injections of ranibizumab, bevacizumab or aflibercept in patients with neovascular AMD. Brit J Ophthalmol. 2014;98(12):1636. doi:10.1136/bjophthalmol-2014-305252
9. Avery RL, Castellarin AA, Steinle NC, et al. SYSTEMIC PHARMACOKINETICS AND PHARMACODYNAMICS OF INTRAVITREAL AFLIBERCEPT, BEVACIZUMAB, AND RANIBIZUMAB. Retin. 2017;37(10):1847-1858. doi:10.1097/iae.0000000000001493
10. Cheungpasitporn W, Chebib FT, Cornell LD, et al. Intravitreal Antivascular Endothelial Growth Factor Therapy May Induce Proteinuria and Antibody Mediated Injury in Renal Allografts. Transplantation. 2015;99(11):2382-2386. doi:10.1097/tp.0000000000000750
11. Morales E, Moliz C, Gutierrez E. Renal damage associated to intravitreal administration of ranibizumab. Nefrología Engl Ed. 2017;37(6):653-655. doi:10.1016/j.nefroe.2017.10.007
12. Shye M, Hanna RM, Patel SS, et al. Worsening proteinuria and renal function after intravitreal vascular endothelial growth factor blockade for diabetic proliferative retinopathy. Clin Kidney J. 2020;13(6):sfaa049-. doi:10.1093/ckj/sfaa049
13. Kameda Y, Babazono T, Uchigata Y, Kitano S. Renal function after intravitreal administration of vascular endothelial growth factor inhibitors in patients with diabetes and chronic kidney disease. J Diabetes Invest. 2018;9(4):937-939. doi:10.1111/jdi.12771
14. O’Neill RA, Gallagher P, Douglas T, et al. Evaluation of long-term intravitreal anti-vascular endothelial growth factor injections on renal function in patients with and without diabetic kidney disease. Bmc Nephrol. 2019;20(1):478. doi:10.1186/s12882-019-1650-1
15. Glassman AR, Liu D, Jampol LM, Sun JK, Network DRCR. Changes in Blood Pressure and Urine Albumin-Creatinine Ratio in a Randomized Clinical Trial Comparing Aflibercept, Bevacizumab, and Ranibizumab for Diabetic Macular Edema. Invest Ophth Vis Sci. 2018;59(3):1199-1205. doi:10.1167/iovs.17-22853
16. Flaxel CJ, Adelman RA, Bailey ST, et al. Age-Related Macular Degeneration Preferred Practice Pattern®. Ophthalmology. 2020;127(1):P1-P65. doi:10.1016/j.ophtha.2019.09.024
17. Flaxel CJ, Adelman RA, Bailey ST, et al. Diabetic Retinopathy Preferred Practice Pattern®. Ophthalmology. 2020;127(1):P66-P145. doi:10.1016/j.ophtha.2019.09.025
18. Network DRCR, Wells JA, Glassman AR, et al. Aflibercept, Bevacizumab, or Ranibizumab for Diabetic Macular Edema. New Engl J Medicine. 2015;372(13):1193-1203. doi:10.1056/nejmoa1414264
19. Flaxel CJ, Adelman RA, Bailey ST, et al. Retinal Vein Occlusions Preferred Practice Pattern®. Ophthalmology. 2020;127(2):P288-P320. doi:10.1016/j.ophtha.2019.09.029
20. Aiello LP, Beck RW, Bressler NM, et al. Rationale for the Diabetic Retinopathy Clinical Research Network Treatment Protocol for Center-Involved Diabetic Macular Edema. Ophthalmology. 2011;118(12):e5-e14. doi:10.1016/j.ophtha.2011.09.058
21. Group CR, Martin DF, Maguire MG, et al. Ranibizumab and Bevacizumab for Neovascular Age-Related Macular Degeneration. New Engl J Medicine. 2011;364(20):1897-1908. doi:10.1056/nejmoa1102673
22. Kertes PJ, Galic IJ, Greve M, et al. Canadian Treat-and-Extend Analysis Trial with Ranibizumab in Patients with Neovascular Age-Related Macular Disease: One-Year Results of the Randomized Canadian Treat-and-Extend Analysis Trial with Ranibizumab Study. Ophthalmology. 2019;126(6):841-848. doi:10.1016/j.ophtha.2019.01.013
23. Norton JM, Ali K, Jurkovitz CT, et al. Development and Validation of a Pragmatic Electronic Phenotype for CKD. Clin J Am Soc Nephro. 2019;14(9):1306-1314. doi:10.2215/cjn.00360119
24. Tian Y, Schuemie MJ, Suchard MA. Evaluating large-scale propensity score performance through real-world and synthetic data experiments. Int J Epidemiol. 2018;47(6):2005-2014. doi:10.1093/ije/dyy120
25. Perazella MA. Pharmacology behind Common Drug Nephrotoxicities. Clin J Am Soc Nephro. 2018;13(12):1897-1908. doi:10.2215/cjn.00150118
26. Wells JA, Glassman AR, Ayala AR, et al. Aflibercept, Bevacizumab, or Ranibizumab for Diabetic Macular Edema Two-Year Results from a Comparative Effectiveness Randomized Clinical Trial. Ophthalmology. 2016;123(6):1351-1359. doi:10.1016/j.ophtha.2016.02.022
27. Gross JG, Glassman AR, Liu D, et al. Five-Year Outcomes of Panretinal Photocoagulation vs Intravitreous Ranibizumab for Proliferative Diabetic Retinopathy: A Randomized Clinical Trial. Jama Ophthalmol. 2018;136(10):1138. doi:10.1001/jamaophthalmol.2018.3255
28. Group C of A related MDTT (CATT) R, Committee: W, Martin DF, et al. Ranibizumab and Bevacizumab for Treatment of Neovascular Age-related Macular Degeneration Two-Year Results. Ophthalmology. 2012;119(7):1388-1398. doi:10.1016/j.ophtha.2012.03.053
29. Cao X, Sanchez JC, Dinabandhu A, et al. Aqueous proteins help predict the response of neovascular age-related macular degeneration patients to anti-VEGF therapy. J Clin Invest. Published online 2021. doi:10.1172/jci144469
30. Heier JS, Campochiaro PA, Yau L, et al. Ranibizumab for Macular Edema Due to Retinal Vein Occlusions Long-term Follow-up in the HORIZON Trial. Ophthalmology. 2012;119(4):802-809. doi:10.1016/j.ophtha.2011.12.005
31. Cox DR. Regression Models and Life‐Tables. J Royal Statistical Soc Ser B Methodol. 1972;34(2):187-202. doi:10.1111/j.2517-6161.1972.tb00899.x



<div id="refs"></div>

\clearpage

\centerline{\Huge Appendix}

# (APPENDIX) Appendix {-}

# Exposure Cohort Definitions

```{r appendix,  echo=FALSE, results="asis", warning=FALSE, message=FALSE}
# source("")

# printCohortDefinitionFromNameAndJson(name = "",
#                                     json = SqlRender::readSql(""),
#                                     withConcepts = TRUE)
```

# Outcome Cohort Definitions

# Negative Control Concepts {#negative-controls}
