---
title: "| RESEARCH PROTOCOL\n| \n| Estimating the comparative risk of kidney failure associated with intravitreal anti-vascular endothelial growth factor exposure in patients with blinding diseases \n"
fontsize: 12pt
geometry: margin=1in
output:
  bookdown::html_document2:
    df_print: paged
    toc: yes
    toc_depth: 2
    toc_float: yes
    number_sections: yes
    number_tables: yes
    css: style.css
  word_document:
    reference_docx: ohdsi-protocol-style.docx
  bookdown::pdf_document2:
    keep_tex: yes
    latex_engine: xelatex
    md_extensions: +raw_attribute
    number_sections: yes
    includes:
      before_body: title.tex
header-includes:
- \usepackage[numbers,sort&compress]{natbib}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{caption}
- \usepackage{rotating}
- \usepackage{multirow}
- \usepackage{mwe,tikz}
- \usepackage[percent]{overpic}
- \usepackage{enumitem}
- \usepackage{hyperref}
- \newcolumntype{P}[1]{>{\raggedright\arraybackslash}p{#1}}
- \newcommand{\footerDate}{`r params$date`}
- \input{header.tex}
longtable: yes
mainfont: Arial
bibliography: Protocol.bib
params:
  date: '28-Mar-2023'
  version: 0.0.1
subtitle: 'Version: `r params$version`'
link-citations: yes
csl: bmj.csl
---

```{r setup, include=FALSE}
library(LegendT2dm)
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
options(kableExtra.latex.load_packages = FALSE)
library(kableExtra)
library(dplyr)
options(knitr.kable.NA = "")
options(knitr.table.format = function() {
  if (knitr::is_latex_output()) {
    "latex"
  } else if (knitr::is_html_output()) {
    "html"
  } else {
    "pipe"
  }
})

latex_table_font_size <- 8
```

# List of Abbreviations

```{r abbreviations, echo=FALSE}
abbreviations <- readr::read_delim(col_names = FALSE, delim = ",", trim_ws = TRUE, file = "
Anti-VEGF, anti-vascular endothelial growth factor 
DME, diabetic macular edema 
DR, diabetic retinopathy 
AMD, age-related macular degeneration 
VO, vein occlusion 
CKD, chronic kidney disease 
ICD, International Classification of Diseases 
CPT, Current Procedural Terminology 
OHDSI, Observational Health Data Sciences and Informatics 
OMOP, Observational Medical Outcomes Partnership 
CDM, common data model 
PS, propensity score 
EHR, electronic health record 
IRB, Institutional Review Board 
JHM, Johns Hopkins Medicine 
ARVO, Association for Research in Vision and Ophthalmology 
")

tab <- kable(abbreviations, col.names = NULL, linesep = "", booktabs = TRUE)

if (knitr::is_latex_output()) {
  tab %>% kable_styling(latex_options = c("striped", "hold_position"), 
                        font_size = latex_table_font_size)
} else {
  tab %>% kable_styling(bootstrap_options = "striped")
}
```

\clearpage

# Responsible Parties

## Investigators

```{r parties, echo=FALSE}
parties <- readr::read_delim(col_names = TRUE, delim = ";", trim_ws = TRUE, file = "
  Investigator; Institution/Affiliation
  Cindy Cai, MD *; Wilmer Eye Institute, Johns Hopkins Hospital, Baltimore, MD, USA
")

tab <- kable(parties, booktabs = TRUE, linesep = "") %>%
  column_spec(1, width = "10em") %>%
  column_spec(2, width = "35em") %>%
  footnote(general = "* Principal Investigator", general_title = "")

if (knitr::is_latex_output()) {
  tab %>% kable_styling(latex_options = c("striped", "hold_position"),
                        font_size = latex_table_font_size)
} else {
  tab %>% kable_styling(bootstrap_options = "striped")
}
```

## Disclosures

This study is undertaken within Observational Health Data Sciences and Informatics (OHDSI), an open collaboration.
**JB** is a fictional character.

\clearpage

# Abstract

Background: The potential adverse kidney effects of the commonly used intravitreal anti-vascular endothelial growth factor (anti-VEGF) medications are poorly understood.
Purpose: To estimate the risk of kidney failure after intravitreal anti-VEGF medications across the Observational Health Data Sciences and Informatics (OHDSI) network, comparing ranibizumab, aflibercept, and bevacizumab.
Design: Retrospective analysis of administrative claims and electronic health record (EHR) data in the OHDSI network.
Subjects: Patients with blinding diseases including diabetic macular edema (DME), diabetic retinopathy (DR), age-related macular degeneration (AMD), and vein occlusion (VO) who receive ≥3 monthly anti-VEGF will be included. Patients with pre-existing kidney failure will be excluded.
Methods: Patients receiving ranibizumab will be compared to propensity score matched patients receiving aflibercept, and patients receiving bevacizumab. Propensity score models will be constructed on baseline characteristics such as demographics (age, sex, race/ethnicity), comorbidities (Charlson Comorbidity Index), and receipt of any nephrotoxic medications during the observation window. Cox models will be used to compare the time to kidney failure between patients who receive ranibizumab with patients who receive aflibercept, ranibizumab with bevacizumab, and bevacizumab with aflibercept.
Conclusions: Knowing the relative risk for kidney failure between the three commonly used anti-VEGF medications will allow ophthalmologists to select safer treatments.


\clearpage

# Amendments and Updates

```{r amendments, echo=FALSE}
amendments <- readr::read_delim(col_names = TRUE, delim = ";", trim_ws = TRUE, file = "
  Number; Date; Section of study protocol; Amendment or update; Reason
")

tab <- kable(amendments, booktabs = TRUE, linesep = "")

if (knitr::is_latex_output()) {
  tab %>% kable_styling(latex_options = c("striped", "hold_position"), font_size = latex_table_font_size)
} else {
  tab %>% kable_styling(bootstrap_options = "striped")
}
```

# Milestones

```{r dates, echo=FALSE}
dates <- readr::read_delim(col_names = TRUE, delim = ";", trim_ws = TRUE, file = "
•	Identify data partners
•	Data partners to share data diagnostics results
•	Complete candidate phenotypes
•	Data partners to share phenotype diagnostics results
•	Finalize analytic package
•	Data partners to run analytic package
•	Data partners to share study diagnostic results
•	Finalize results from data partners
•	Manuscript drafting and writing
•	Submit abstract: OHDSI Global Symposium (May 2023), ARVO (Jan 2024)
")

tab <- kable(dates, booktabs = TRUE, linesep = "")

if (knitr::is_latex_output()) {
  tab %>% kable_styling(latex_options = c("striped", "hold_position"), font_size = latex_table_font_size)
} else {
  tab %>% kable_styling(bootstrap_options = "striped")
}
```

# Rationale and Background
Intravitreal anti-vascular endothelial growth factor (anti-VEGF) medications have revolutionized the treatment of blinding diseases.1,2 Nearly 20 million intravitreal injections were given globally in 2016 to treat conditions such as diabetic macular edema (DME), diabetic retinopathy (DR), age-related macular degeneration (AMD), and vein occlusion (VO).1 When given systemically, anti-VEGF medications have adverse kidney effects (e.g., proteinuria, acute kidney injury).3–5 Patients at high risk for chronic kidney disease (CKD), for example, those ≥60 years and those with diabetes, may be preferentially harmed by the adverse kidney effects of intravitreal anti-VEGF.6 CKD can progress to kidney failure, which is a major source of morbidity and mortality for the patient and is a financial burden to society.6,7

When anti-VEGF medications are given intravitreally, there is systemic absorption of the drug that is greatest with bevacizumb and aflibercept and lowest with ranibizumab.8,9 However, evidence about the subsequent kidney effect is conflicting. Several case reports document kidney damage after intravitreal anti-VEGF.10–12 On the other hand, retrospective cohort studies have not identified treatment-associated declines in kidney function, or estimated glomerular filtration rate (eGFR).13,14 A secondary analysis from a randomized trial also did not show differences in rates of kidney damage.15 However, these studies were either limited to a single institution, had small sample sizes (<600 patients), only included patients without CKD, and did not evaluate the severe phenotype of kidney failure. To resolve this controversy, a comprehensive study is needed to evaluate the kidney effects of intravitreal anti-VEGF. In populations at high risk for CKD, the treating ophthalmologist can choose alternate forms of therapy (e.g., focal laser for DME), or the intravitreal anti-VEGF with the least systemic absorption (e.g., ranibizumab).16–18


# Study Objectives
The overall objective of the proposal is to estimate the comparative risk of kidney failure associated with intravitreal anti-vascular endothelial growth factor exposure in patients with blinding diseases. The rationale is that knowing the comparative risk for kidney failure among the three commonly used anti-VEGF medications will allow the ophthalmologist to select the safest treatments, particularly for populations at high risk for adverse kidney effects. To address our hypothesis, we propose the following specific aims.
1.	Amongst people with blinding diseases (DR/DME, AMD, and VO), does exposure to ranibizumab increase the risk of kidney failure, relative to aflibercept?
2.	Amongst people with blinding diseases (DR/DME, AMD, and VO), does exposure to ranibizumab increase the risk of kidney failure, relative to bevacizumab?
3.	Amongst people with blinding diseases (DR/DME, AMD, and VO), does exposure to bevacizumab increase the risk of kidney failure, relative to aflibercept?

# Research Methods

## Study Design


## Data Sources

## Study Population

## Exposure Comparators

## Outcomes {#outcomes}

## Analysis

# Sample Size and Study Power {#sample-size}

# Strengths and Limitations {#strengths-limitations}

# Protection of Human Subjects

# Management and Reporting of Adverse Events and Adverse Reactions

# Plans for Disseminating and Communicating Study Results

\clearpage

# References {-}

<div id="refs"></div>

\clearpage

\centerline{\Huge Appendix}

# (APPENDIX) Appendix {-}

# Exposure Cohort Definitions

```{r appendix,  echo=FALSE, results="asis", warning=FALSE, message=FALSE}
source("")

printCohortDefinitionFromNameAndJson(name = "",
                                     json = SqlRender::readSql(""),
                                     withConcepts = TRUE)
```

# Outcome Cohort Definitions

# Negative Control Concepts {#negative-controls}
